---
title: "Forecasting with odin2 models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting with odin2 models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Overview

This vignette shows how to use an SIR model written with `odin2` to produce an
`incast` forecast.

`odin2` describes the dynamical model. We use it to write the SIR equations and
the comparison between modelled incidence and observed data. It does not fit
the model.

`dust2` runs the odin2 model for a given set of parameters. It compares the
model output with the observations and calculates the likelihood: how well
those parameters explain the data.

`monty` performs the inference. It combines the likelihood from `dust2` with the
prior and uses Markov chain Monte Carlo (MCMC) to draw parameter values from
the posterior.


`incast.odin` connects these pieces to the `incast` pipeline.

A custom `incast.odin::ODIN2()` model requires two inputs:

1. the model from `odin2::odin()`;
2. the prior from `monty::monty_dsl()`.

You do not need to write `dust2` or MCMC code. The current interface uses a deterministic epidemic trajectory. Forecast uncertainty comes from posterior parameter draws and the observation model.

## Example: one New York influenza wave

This example uses New York influenza hospital admissions from the 2024–25
season. The `as_of` filter recreates the data available on 23 February 2025;
later revisions and future observations are excluded.

```{r packages-data}
library(incast)
library(incast.odin)
library(odin2)
library(monty)
library(dplyr)

snapshot <- as.Date("2025-02-23")

ny_flu <- example_data |>
  filter(
    location == "NY",
    target_end_date >= as.Date("2024-10-19"),
    as_of <= snapshot
  )
```

A closed SIR model should not be fitted across several influenza seasons because it contains no seasonal forcing, importation or replenishment of susceptible people.

## Model structure

The SIR model contains three compartments and two flows:

| Symbol | Meaning |
|---|---|
| `S` | Susceptible people |
| `I` | Infectious people |
| `R` | Recovered people |
| `infections` | Flow from `S` to `I` |
| `recoveries` | Flow from `I` to `R` |

`incidence` accumulates new infections during one reporting interval. Because
the input data are weekly, one model-time unit is one week. The default
`dt = 0.25` divides each week into four model updates.

```{r generator}
sir <- odin2::odin({
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0

  update(S) <- S - infections
  update(I) <- I + infections - recoveries
  update(R) <- R + recoveries
  update(incidence) <- incidence + infections

  infection_probability <- 1 - exp(-beta * max(I, 0) / N * dt)
  recovery_probability <- 1 - exp(-gamma * dt)
  infections <- max(S, 0) * infection_probability
  recoveries <- max(I, 0) * recovery_probability

  beta <- parameter()
  gamma <- parameter()
  N <- parameter()
  I0 <- parameter()
  rho <- parameter()
  phi <- parameter()

  cases <- data()
  cases ~ NegativeBinomial(
    size = phi,
    mu = rho * incidence + 1e-6
  )
})
```

The final two statements tell `dust2` how to evaluate observed hospital
admissions against the latent infections:

- `cases <- data()` reads the observed counts;
- `cases ~ NegativeBinomial(...)` defines their likelihood.

## Parameters and prior

| Parameter | Meaning |
|---|---|
| `beta` | Weekly transmission rate |
| `gamma` | Weekly recovery rate |
| `I0` | Infectious people at the start of the window |
| `rho` | Probability that an infection appears in the observed hospital target |
| `phi` | Negative binomial overdispersion |
| `N` | Population, fixed here at 19.5 million |

Parameters named in the monty prior are estimated. Fixed parameters are
passed separately. These priors are working assumptions for this example and
should be reviewed for a real analysis.

```{r prior}
prior <- monty::monty_dsl({
  beta ~ LogNormal(meanlog = log(2.5), sdlog = 0.3)
  gamma ~ LogNormal(meanlog = log(1.75), sdlog = 0.25)
  rho ~ Beta(a = 2, b = 398)
  I0 ~ LogNormal(meanlog = log(1000), sdlog = 0.8)
  phi ~ Exponential(mean = 30)
})

fixed <- list(N = 19.5e6)
```

## Create the incast model

`ODIN2()` combines the generator, prior and fixed parameters into a model that
`incast` can fit and forecast.

```{r adapter}
set.seed(1)

sir_model <- ODIN2(
  observation,
  generator = sir,
  prior = prior,
  fixed = fixed,
  observation_model = odin_negbin(size = "phi", scale = "rho"),
  n_steps = 5000,
  n_draws = 500
)
```

`n_steps` controls the parameter sampling. `n_draws` is the number of
posterior trajectories retained for forecasting.

There are two negative binomial statements because they serve different
directions of the workflow:

| Statement | Purpose |
|---|---|
| `cases ~ NegativeBinomial(...)` inside odin2 | Score historical observations while fitting |
| `odin_negbin()` passed to `ODIN2()` | Simulate future observations after projecting incidence |

They must describe the same observation process.

The adapter requires only two naming conventions in the odin2 model:

- observed counts are named `cases`;
- the scalar forecast state is named `incidence`.

## Compare and forecast with incast

Once created, `sir_model` behaves like the models returned by
`default_models()`. `get_cv()` refits it at every forecast origin.
`get_fcast()` then refits the selected model to all available observations and
projects it into the forecast period.

```{r pipeline}
models <- c(
  default_models(),
  list(SIR = sir_model)
)

ncast <- ny_flu |>
  check_data() |>
  get_ncast(max_delay = 2, draws = 200)

cv <- ncast |>
  get_cv(
    h = 2,
    n_origins = 4,
    step = 1,
    models = models
  )

cv$score
autoplot(cv)

fcast <- cv |>
  get_fcast(top_n = 1)

autoplot(fcast)
```

## Built-in SIR shortcut

The package includes the same SIR generator already compiled:

```{r shortcut}
sir_model <- odin_sir(
  observation,
  population = 19.5e6,
  prior = prior,
  n_steps = 5000,
  n_draws = 500
)
```

Use `odin_sir()` to start from the supplied equations. Use `ODIN2()` when
changing the model, likelihood or future observation process. The downstream
`incast` pipeline is identical.

## Further reading

- `vignette("odin2", package = "odin2")` introduces the modelling language.
- `vignette("fitting", package = "odin2")` develops the dust2 and monty fitting
  workflow in detail.
- The [odin2 documentation](https://mrc-ide.github.io/odin2/) covers model
  syntax, fitting and packaging.
